---
title: "README"
author: ""
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("Packages.R")
```

This folder contains the necessary code to compute the various validation results for the updated PANGEA trajectory models (BM and no BM) and alternatives (original PANGEA and 20/2/20).

The folder includes:
\begin{itemize}
\item validation\_code.R: main script to run to perform the analysis
\item data\_preparation.R: script computing the trajectories, risk scores, and some other variables needed for the analysis
\item table\_descriptives.R: script to compute the basic descriptive statistics of the cohort (sourced in validation\_code.R)
\item risk\_reclassification.R: script to compute risk strata from the PANGEA risk scores (sourced in validation\_code.R)
\item Functions.R: script defining the functions needed to run the various analyses
\item Packages.R: script loading the needed packages
\item cutoffs\_pangea\_training.rda: file storing cutoffs used to determine risk strata
\item models: folder storing the PANGEA models and necessary information to compute the risk predictions
\item results: folder to save the results at the end of the analysis (currently empty)
\end{itemize}

\subsection{How to perform the analysis}

\begin{itemize}
\item Set the "Validation - Code for External Collaborators" folder as working directory
\item Open the R script "validation\_code.R"
\item Modify the path to load your data (at the beginning of the script)
\item If applicable, on line 7 set save\_deidentified to TRUE (instead of FALSE, the default value)
\item Run the "validation\_code.R" script
\end{itemize}
The results will be saved in the results folder.

The validation_code.R executes the following steps:
\begin{enumerate}
\item Source the Functions.R and Packages.R scripts
\item Load the data
\item Source the data\_preparation.R script to compute some variables needed for the analysis, including the trajectories, the rolling 20/2/20 classification, and PANGEA risk scores (including for some previous developmental versions of the model)
\begin{enumerate}
\item \textbf{Note}: To just compute the PANGEA 2.0 risk scores, simply run validation\_code.R up to line 32 where data\_preparation.R is sourced.
\item The PANGEA 2.0 models are saved in the pangea\_bm\_traj\_alt.RData and pangea\_no\_bm\_traj\_alt.RData files in the models folder, and their risk score variables created by data\_preparation.R have similar names.
\end{enumerate}
\item Compute descriptive statistics for the cohort
\item Compute basic statistics on progression rates for the cohort
\item Compute descriptive statistics on the risk scores for the cohort
\item Compute concordance statistics (ability of the risk scores to appropriately rank patients)
\item Compute classification statistics (based on risk strata)
\item Compute calibration statistics (ability of risk scores to match the actual progression risk for various subgroups)
\item Store the results in the results folder
\end{enumerate}


\subsection{Format of the data}

The dataset has to include the following variables:
\begin{description}
\item[tstart:] Time at which the current labs are measured (it should be 0 for the first observation) in years; used for validation.
\item[tstop:] Time of the next lab or the end of the follow up (final observation) in years; used for validation.
\item[prog\_mm:] Binary progression status at the time tstop (1 = slim-crab myeloma, 0 otherwise); used for validation.
\item[mspike:] M-spike in g/dL; used for the PANGEA risk prediction.
\item[iuratio:] Involved/Uninvolved free light chain ratio; used for the PANGEA risk prediction.
\item[plasmacells:] Numeric percentage of plasma cells (0-100); used for the PANGEA risk prediction.
\item[creatinine:] Creatinine in mg/dL; used for the PANGEA risk prediction.
\item[age:] Age of the patient at each visit in years; used for the PANGEA risk prediction.
\item[hgb:] Hemoglobin in g/dL (needed to compute the trajectory); used for the PANGEA risk prediction.
\item[sex:] Sex of the patient; used only for descriptive statistics.
\item[current\_diagnosis:] Categorical variable, is the patient smoldering, i.e. "SMM" (for this analysis we want only smoldering patients); used only for descriptive statistics.
\item[end\_date\_type:] Description of the event that caused censoring or progression. possible values for censored patients are "Death", "Last appointment", and "Other treatment". For patients who progress the value should just be "Diagnosis". Used only for descriptive statistics.
\item[race:] Race of patient. Possible values are "Asian", "Black or African American", "Declined", "Multiple", "Other", and "White". For patients where race is missing (it's ok if that's everyone in your cohort), the value should be NA. Used only for descriptive statistics.
\item[ethnicity:] Ethnicity of the patient. Possible values are "Declined", "Hispanic or Latino", and "Not Hispanic or Latino". For patients where ethnicity is missing (it's ok if that's everyone in your cohort), the value should be NA. Used only for descriptive statistics.
\item[immunofix2:] Immunofixation isotype. Possible values are "Biclonal", "IgA", "IgG", and "Light Chain Only". Missing values should be indicated as NA. Used only for descriptive statistics.
\end{description}

The data has to be in a long format, which is, each patient visit corresponds to
one row in the dataset. 

\textbf{Note}: PANGEA risk predictions can only be computed for the rows in which all predictor variables (mspike, iuratio, creatinine, hgb, and age) are available (not NA). Because of this, labs associated with the same patient visit but reported on slightly different dates (e.g. mspike on 5/9/24 and iuratio on 5/4/24) should be combined into a single row with a single date in order to get a risk prediction for that visit.

For a visual aide, please see the first rows of our dataset, 
obtained using the R function \texttt{tmerge} from the package \texttt{survival}. 
The data are loaded in data.table format.

```{r}
## Dataset
library(data.table)
sample_data <- fread("fake_example_data.csv")
sample_data[participant_id == 1,]
sample_data[participant_id == 2,]
```

\section{R session}
As a reference, we report the R version we used

```{r}
R.Version()$version.string
```

